Exploring dimensions of biodiversity in Japanese ferns

### 2021-08-18 ms/SI.Rmd
Add non-ultrametric model

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp",
  "morans_i",
  "t_test_results",
  "aic_env_repro",
  "comm_ferns",
  "repro_data",
  "japan_pterido_tree"
))
```

@@

# Set color palettes
# - Randomization test significance
signif_cols <-
  c(
    "> 0.99" = "#e31a1c", 
    "> 0.975" = "#fb9a99",
    "< 0.01" = "#1f78b4", 
    "< 0.025" = "#a6cee3",
    "not significant" = "#ffffcc"
  )
# - CANAPE
canape_cols <-
  c(
    "neo" = "#e31a1c", 
    "paleo" = "#1f78b4",
    "mixed" = "#6a51a3",
    "super" = "#4a1486",
    "not significant" = "#ffffcc"
  )
# - Bioregions
bioregion_cols <- rcartocolor::carto_pal(9, "Bold")[1:8] %>%
  set_names(1:8)
```

@@

>>>>

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp",
  "morans_i",
  "t_test_results",
  "aic_env_repro",
  "comm_ferns",
  "repro_data",
  "japan_pterido_tree",
  "model_fits_not_ult",
  "biodiv_ferns_cent"
))
```

@@

# Set color palettes
# - Randomization test significance
signif_cols <-
  c(
    "> 0.99" = "#e31a1c", 
    "> 0.975" = "#fb9a99",
    "< 0.01" = "#1f78b4", 
    "< 0.025" = "#a6cee3",
    "not significant" = "#ffffcc"
  )
# - CANAPE
canape_cols <-
  c(
    "neo" = "#e31a1c", 
    "paleo" = "#1f78b4",
    "mixed" = "#6a51a3",
    "super" = "#4a1486",
    "not significant" = "#ffffcc"
  )
# - Bioregions
bioregion_cols <- rcartocolor::carto_pal(9, "Bold")[1:8] %>%
  set_names(1:8)

# Specify subplot order
model_subplot_order <- tribble(
  ~resp_var, ~indep_var, ~order,
  "richness", "temp", 1,
  "pd_obs_z", "temp", 2,
  "rpd_obs_z", "temp", 3,
  "fd_obs_z", "temp", 4,
  "rfd_obs_z", "temp", 5,
  "pd_obs_z", "percent_apo", 6,
  "rpd_obs_z", "percent_apo", 7
)
```

@@

```{r modelfit-not-ult, fig.width = 7, fig.height = 3}
# Extract dataset used for model, including significance columns
biodiv_ferns_used_for_model <- biodiv_ferns_spatial %>%
  # drop geometry (only need centroids for modeling)
  st_set_geometry(NULL) %>%
  as_tibble() %>%
  # filter to only sites used in model
  inner_join(
    biodiv_ferns_cent %>% as_tibble() %>% select(grids),
    by = "grids"
  )

# Generate final plot
modelfit_plot_list <- 
  model_fits_not_ult %>%
  mutate(
    signif_var = str_replace_all(resp_var, "_obs_z", "_signif"),
    resp_var_print = case_when(
      resp_var == "richness" ~ "Richness (no. spp.)",
      TRUE ~ str_to_upper(resp_var) %>% word(1, 1, "_") %>% paste("SES of", .)
    ),
    model_data = list(biodiv_ferns_used_for_model)) %>%
  left_join(model_subplot_order, by = c("resp_var", "indep_var")) %>%
  arrange(order) %>%
  pmap(plot_fits)

# Need one subplot with a legend
modelfit_plot_list[[2]] <- modelfit_plot_list[[2]] +
  theme(legend.position = "right") +
  labs(color = "Significance") +
  # Increase size of points in legend
  guides(colour = guide_legend(override.aes = list(size = 1.5)))

wrap_plots(modelfit_plot_list) +
  guide_area() +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") 
```

**`r figure("modelfit-not-ult")`**. Relationship between (**A**) phylogenetic diversity (PD) and (**B**) relative PD and % apomictic taxa, models inferred using phylogeny with branchlengths in units of expected genetic change (untransformed ML tree). Ribbon shows 95% confidence interval of model. Line fit with the focal predictor variable while averaging over other predictors. For RPD, the standard effect size (SES) was calculated by comparing raw values to a null distribution of 999 random values, and significance (*P*-value) calculated using a two-tailed test (see Materials and Methods).

### 2021-08-18 ms/manuscript.Rmd
Add non-ultrametric model

Furthermore, we identified % apomictic taxa as a strong driver of SES of PD (`r figure("modelfit")`G).
Indeed, *Dryopteris* is one such genus with a high rate of apomictic taxa (`r s_figure("apo-genus")`), and other genera with high rates of apomixis (*Cyrtomium*, *Pteris*; `r s_figure("apo-genus")`) also reach greatest richness in southern Honshu (`r s_figure("genus-map")`).
This is consistent with the hypothesis that communities with high proportions of asexual taxa have lower genetic diversity, and is the first time to our knowledge that this has been demonstrated in ferns. 
While temperature is also correlated with % apomictic taxa and therefore cannot be discounted, conditional AIC indicates a better fit of the reproductive model than the environmental model for SES of RPD (`r s_table("aic")`). 
Ultimately, it is likely that temperature, in combination with other environmental conditions, is driving the predominance of apomictic taxa: @Tanaka2014 also found that the proportion of apomictic taxa reaches a maximum at ca. 34.8 °N latitude and increases with temperature in Japan (their study was limited to Honshu and did not include the southern islands). 
@Tanaka2014 asserted that one limiting factor affecting establishment of apomictic ferns at high latitudes or cold climates is the large cell size of apomicts, which are typically polyploid. 
As fern gametophytes are only a single cell-layer in thickness with no cuticle or stomata, it is expected they would be particularly susceptible to freezing damage induced by larger cells with higher water content; however, no experimental studies have tested these hypotheses.

@@

# Define function for plotting model fit and actual data
plot_fits <- function(resp_var, resp_var_print, indep_var, signif_var, fit, model_data, ...) {

  # Setup randomization test significance colors
  signif_cols <-
    c(
      "> 0.99" = "#e31a1c", 
      "> 0.975" = "#fb9a99",
      "< 0.01" = "#1f78b4", 
      "< 0.025" = "#a6cee3",
      "not significant" = "#ffffcc"
    )

  # Basic plot setup
  p <- ggplot(model_data, aes(x = .data[[indep_var]], y = .data[[resp_var]])) +
    labs(y = resp_var_print) +
    theme_gray(base_size = 9) +
    theme(panel.grid.minor = element_blank())

  # Add points, color by significance if not plotting richness
  if (resp_var != "richness") {
    p <- p + 
      geom_point(aes(color = .data[[signif_var]]), size = 0.5) +
      scale_color_manual(values = signif_cols)
  } else {
    p <- p + 
      geom_point(size = 0.5, color = "grey20")
  }

  # Add line and ribbon of model fit
  p <- p +
    geom_line(data = fit, alpha = 0.35, color = "blue") +
    geom_ribbon(
      data = fit,
      aes(ymin = low, ymax = up), 
      alpha = 0.15
    ) +
    theme(legend.position = "none")

  # Format x-axis ticks and labels
  if (indep_var == "temp") {
    p <- p +
      scale_x_continuous(labels = function(x) x*0.1) +
      labs(x = "Temperature (°C)")
  } else if (indep_var == "percent_apo") {
    p <- p +
      scale_x_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
      labs(x = "% apomictic taxa")
  }

  p

}

>>>>

Furthermore, we identified % apomictic taxa as a strong driver of SES of RPD (`r figure("modelfit")`G).
Indeed, *Dryopteris* is one such genus with a high rate of apomictic taxa (`r s_figure("apo-genus")`), and other genera with high rates of apomixis (*Cyrtomium*, *Pteris*; `r s_figure("apo-genus")`) also reach greatest richness in southern Honshu (`r s_figure("genus-map")`).
This is consistent with the hypothesis that communities with high proportions of asexual taxa have lower genetic diversity, and is the first time to our knowledge that this has been demonstrated in ferns. 
We verified that this relationship is also observed when analyzed with a phylogeny with branch lengths in units of genetic change (`r ie`, the output of the maximum likelihood phylogenetic analysis, before dating), consistent with an effect on genetic diversity (`r s_figure("modelfit-not-ult")`B).
While temperature is also correlated with % apomictic taxa and therefore cannot be discounted, conditional AIC indicates a better fit of the reproductive model than the environmental model for SES of RPD (`r s_table("aic")`). 
Ultimately, it is likely that temperature, in combination with other environmental conditions, is driving the predominance of apomictic taxa: @Tanaka2014 also found that the proportion of apomictic taxa reaches a maximum at ca. 34.8 °N latitude and increases with temperature in Japan (their study was limited to Honshu and did not include the southern islands). 
@Tanaka2014 asserted that one limiting factor affecting establishment of apomictic ferns at high latitudes or cold climates is the large cell size of apomicts, which are typically polyploid. 
As fern gametophytes are only a single cell-layer in thickness with no cuticle or stomata, it is expected they would be particularly susceptible to freezing damage induced by larger cells with higher water content; however, no experimental studies have tested these hypotheses.

@@
